Approximation as Differentiation

### 2025-04-16 appendix/notes/cmon-enriched.tex
Standardise on X,Y,Z,W for objects.

\begin{definition}[Commutative monoid]
A \emph{commutative monoid} $A = (A, \varepsilon, \bullet)$ is a set $A$ equipped with distinguished element
$\varepsilon \in A$ called the \emph{unit} and associative binary operation $\bullet$ satisfying $\varepsilon
\bullet x = x$ and $x \bullet \varepsilon = x$ for any $x \in A$.
\end{definition}

A commutative monoid homomorphism from $A$ to $B$ is any function $f: A \to B$ preserving $\varepsilon$ and
$\bullet$.

\subsubsection{Examples of commutative monoids}

\begin{definition}[Bounded semilattice]
\label{def:cmon-enriched:bounded-semilattice}
A \emph{bounded semilattice} $A = (A, \varepsilon, \bullet)$ is a commutative monoid where $\bullet$ is
\emph{idempotent}, i.e.~satisfies $x \bullet x = x$.
\end{definition}

\noindent The idempotence of $\bullet$ (together with commutativity and associativity) induces a partial order
$\le_\bullet$ on $X$, with $x \le_{\bullet} y \iff x \bullet y = x$. With respect to this partial order, $x
\bullet y$ is the greatest lower bound (meet) of $x$ and $y$ and $\varepsilon$ is the top element; with
respect to the opposite order, $x \bullet y$ is the least upper bound (join) of $x$ and $y$ and $\varepsilon$
is the bottom element. This therefore provides an algebraic characterisation of the usual (dual)
order-theoretic notions of bounded meet semilattices $(X, \top, \meet)$ and bounded join semilattices $(X,
\bot, \join)$.

@@

\begin{definition}[Category $\CMon$]
Define $\CMon$ to be the category whose objects $A$ are the commutative monoids $A$ and morphisms $f: A \to B$
are the commutative monoid homomorphisms.
\end{definition}

@@

@@

Recall that if a category $\cat{C}$ is $\CMon$-enriched, then:
\begin{enumerate}
\item Every hom-object $\Hom{\cat{C}}{A}{B}$ is a commutative monoid of morphisms; we write $\zero_{A,B}$
(zero morphism) for the unit, and $+_{A,B}$ (addition of morphisms) for the binary operation.
\item Composition is \emph{bilinear}, i.e.~given by a family of morphisms $\Hom{\cat{C}}{B}{C} \tensor
\Hom{\cat{C}}{A}{B} \to \Hom{\cat{C}}{A}{C}$ in $\CMon$ that preserve the additive structure in
$\Hom{\cat{C}}{B}{C}$ and $\Hom{\cat{C}}{A}{B}$ separately:

@@

Because $\CMon$ is monoidal closed, it is enriched over itself; every hom-object $\Hom{\CMon}{A}{B}$ is a
commutative monoid where:

\begin{enumerate}
\item the unit $0$ is the constant homomorphism sending every element of $A$ to $\varepsilon_B$;
\item the binary operation $+$ is pointwise addition of homomorphisms $(f + g)(a) = f(a) + g(a)$.
\end{enumerate}

>>>>

\begin{definition}[Commutative monoid]
A \emph{commutative monoid} $X = (X, \bullet, \varepsilon)$ is a set $X$ equipped with distinguished element
$\varepsilon \in X$ called the \emph{unit} and associative binary operation $\bullet: X^2 \to X$ satisfying
$\varepsilon \bullet x = x$ and $x \bullet \varepsilon = x$ for any $x \in X$.
\end{definition}

A commutative monoid homomorphism from $X$ to $Y$ is any function $f: X \to Y$ preserving $\varepsilon$ and
$\bullet$.

@@

\begin{definition}[Category $\CMon$]
Define $\CMon$ to be the category which has as objects $X$ all commutative monoids and as morphisms $f: X \to
Y$ all commutative monoid homomorphisms.
\end{definition}

@@

\subsubsection{Example of commutative monoid}

\begin{definition}[Bounded semilattice]
\label{def:cmon-enriched:bounded-semilattice}
A \emph{bounded semilattice} $X = (X, \bullet, \varepsilon)$ is a commutative monoid where $\bullet$ is
\emph{idempotent}, i.e.~satisfies $x \bullet x = x$.
\end{definition}

\noindent The idempotence of $\bullet$ (together with commutativity and associativity) induces a partial order
$\le_\bullet$ on $X$, with $x \le_{\bullet} y \iff x \bullet y = x$. With respect to this partial order, $x
\bullet y$ is the greatest lower bound (meet) of $x$ and $y$ and $\varepsilon$ is the top element; with
respect to the opposite order, $x \bullet y$ is the least upper bound (join) of $x$ and $y$ and $\varepsilon$
is the bottom element. This therefore provides an algebraic characterisation of the usual (dual)
order-theoretic notions of bounded meet semilattices $(X, \meet, \top)$ and bounded join semilattices $(X,
\join, \bot)$.

\begin{definition}[Category $\SemiLat$]
Define $\SemiLat$ to be the category which has as objects all bounded semilattices and as morphisms all
bounded semilattice homomorphisms.
\end{definition}

@@

Recall that if a category $\cat{C}$ is $\CMon$-enriched, then:
\begin{enumerate}
\item Every hom-object $\Hom{\cat{C}}{X}{Y}$ is a commutative monoid of morphisms; we write $\zero_{X,Y}$
(zero morphism) for the unit, and $+_{X,Y}$ (addition of morphisms) for the binary operation (omitting the
indices where implied by the context).
\item Composition is \emph{bilinear}, i.e.~given by a family of morphisms $\Hom{\cat{C}}{Y}{Z} \tensor
\Hom{\cat{C}}{X}{Y} \to \Hom{\cat{C}}{X}{Z}$ in $\CMon$ that preserve the additive structure in
$\Hom{\cat{C}}{Y}{Z}$ and $\Hom{\cat{C}}{X}{Y}$ separately:

@@

Because $\CMon$ is monoidal closed, it is enriched over itself; every hom-object $\Hom{\CMon}{X}{Y}$ is a
commutative monoid with:

\begin{enumerate}
\item unit $0_{X,Y} = \const(\varepsilon_Y)$, the constant homomorphism sending every element of $X$ to
$\varepsilon_Y$;
\item binary operation $+_{X,Y}$ given by pointwise addition of homomorphisms $(f + g)(x) = f(x) + g(x)$.
\end{enumerate}
